\Large \bf Relationship of Weather and Maize Yields in Kenya

### 2018-11-14 writing/draft1/manuscript17.tex
adding

Before we started our research, we had to answer the following questions: `How should we measure food security?' And `How should we measure drought?' To answer the first question, the measure which has been used as a proxy for food security in the literature are as follows:\\

@@

Thus, after conducting a rigorous literature research on measures of drought, we decided to utilize raw daily precipitation data \textcolor{red}{from...CHIRPS?? with resolution... and daily temperature data from..at resolution....}. The frequency of the maize yield data is yearly while the weather data are daily. Hence, the weather data need to be aggregated in order to obtain a dataset conformable with the yield data. There are many possibilities how to aggregate daily weather data. Commonly used measures of precipitation and temperature include monthly averages (or monthly totals in case of precipitation) and their variances or standard deviations (\citealt{AbrahaSavage2006, LobellEtAl2008, ThorntonEtAl2009}). \cite{Adejuwon2004} analysed relationship of crop yields and three measures of precipitation. The measures of precipitation include: \textit{(i)} Total rainfall during the first month of the period from sowing to harvesting (June) \textit{(ii)} Total rainfall during the first two months of the period from sowing to harvesting (June and July) and \textit{(iii)} Total rainfall during the first three months of the period from sowing to harvesting (June, July and August). Based on his results, weather during June and July is the most important for crop yield in Sub-Saharan West Africa \citep{Adejuwon2004}. Other studies utilised seasonal totals or means (\citealt{sagoe2006,LobellBurke2010}) or annual totals or means \citep{BLIGNAUT2009}.  Some studies are based on simulated daily extremes, averages, daily measures of variance  (\citealt{SchulzeEtA1993,Chipanshi2003,AbrahaSavage2006}) or yearly extremes \citep{sagoe2006}. Another measures which have been proposed for modelling the variability of maize yield are numbers of wet and dry days per a defined period, usually a month or a season (\citealt{BenMohamed2002,AbrahaSavage2006,sagoe2006,Giannakopoulos2009}) or length of rainy season (\citealt{Leemans1993,BenMohamed2002}). The definition of wet and dry days and rainy season vary across the literature. For example, \cite{BenMohamed2002} has assumed that rainy season begins when the amount of rainfall in three consecutive days reach at least 25mm and no dry spell of more than seven days occurs in the following thirty days. According to this study, the end of the rainy season is defined as that rainy day after which rain recorded during 20 days is less than 5mm. \cite{BenMohamed2002} has also found sea surface temperature anomalies at various locations and amount of rainfall in July, August and September to be significant for millet crops in Niger. The author has also considered the maximum air temperature in the hottest month (April) and the minimum air temperature in the coldest month (January) as possible predictors of crops in Niger, but he did not find them significant.

An important group of studies that analyse relationship of yield and climate in Sub-Saharan Africa has utilised degree days (\citealt{SchulzeEtA1993,TingemEtAl2008,WalkerSchulze2008,TingemEtAl2009}) or number of days with temperature above certain level or within defined range (\citealt{Giannakopoulos2009,LauxEtAl}).

(\citealt{ThorntonEtAl2009}) \textcolor{red}{also uses diurnal daily}

@@

Blignaut: annual averages
Chipanschi: daily maximum and minimum temperatures
Giannoukopoulos: number of hw days and so...
Laux et al.: daily Tmax >30C
Lobell et al. 2008: Monthly temp and prec.
Lobell and Burke 2010: Temp. growing season average, precip:growing season total

(\citealt{AbrahaSavage2006,Adejuwon2004,BenMohamed2002,BLIGNAUT2009,
Chipanshi2003,Giannakopoulos2009,LauxEtAl,Leemans1993,LobellEtAl2008,
LobellBurke2010,sagoe2006,SchulzeEtA1993,ThorntonEtAl2009,TingemEtAl2008,
TingemEtAl2009,WalkerSchulze2008})

>>>>

Before beginning our research, we had to answer the following two questions: `How should we measure food security?' and `How should we measure drought?' Based on previous literature and data availability, there are three main approaches of how to measure food security. These include agricultural crop yields (see literature overview in Sections~\ref{Introduction} and further in this section), food prices or health or utility indicators. The relevant studies focused on food prices 
were used to estimate relationship between sea surface temperature and commodity prices in tropics

We focused on maize as this crop is the principal staple food in Kenya and it is grown at 90\% of all farms in Kenya \citep{FAO}.

\cite{brown2014book}: global, food prices and so \\
\cite{BROWN2015} Weather and international price shocks on food prices in the developing world.  Markets in Africa, South Asia, Latin America (regions that contain large segments of poor) \\
\cite{OxfamIDS}: GLOBE, CGE sub-Saharan Africa, central America north Africa, other low-income countries selected by OXFAM  \\
\cite{robinson2010}   Implications of food production and price shocks welfare in Ethiopia: a general equilibrium analysis CGE \\
\cite{Ubilava2018}: Estimate relationship between sea surface temperature and commodity prices using. tropics?? not sure about geographic focus entirely...

@@

Thus, after conducting a rigorous literature research on measures of drought, we decided to utilize raw daily precipitation data \textcolor{red}{from...CHIRPS?? with resolution... and daily temperature data from..at resolution....}. The frequency of the maize yield data is yearly while the weather data are daily. Hence, the weather data need to be aggregated in order to obtain a dataset conformable with the yield data. There are many possible ways how to aggregate daily weather data. Commonly used measures of precipitation and temperature include monthly averages (or monthly totals in case of precipitation) and their variances or standard deviations (\citealt{AbrahaSavage2006, LobellEtAl2008, ThorntonEtAl2009}). \cite{Adejuwon2004} analysed relationship of crop yields and three measures of precipitation. The measures of precipitation include: \textit{(i)} Total rainfall during the first month of the period from sowing to harvesting (June) \textit{(ii)} Total rainfall during the first two months of the period from sowing to harvesting (June and July) and \textit{(iii)} Total rainfall during the first three months of the period from sowing to harvesting (June, July and August). Based on his results, weather during June and July is the most important for crop yield in Sub-Saharan West Africa \citep{Adejuwon2004}. Other studies utilised seasonal totals or means (\citealt{sagoe2006,LobellBurke2010}) or annual totals or means \citep{BLIGNAUT2009}.  Some studies are based on simulated daily extremes, averages, daily measures of variance  (\citealt{SchulzeEtA1993,Chipanshi2003,AbrahaSavage2006}) or yearly extremes \citep{sagoe2006}. Another measures which have been proposed for modelling the variability of maize yield are numbers of wet and dry days per a defined period, usually a month or a season (\citealt{BenMohamed2002,AbrahaSavage2006,sagoe2006,Giannakopoulos2009}) or length of rainy season (\citealt{Leemans1993,BenMohamed2002}). The definition of wet and dry days and rainy season vary across the literature. For example, \cite{BenMohamed2002} has assumed that rainy season begins when the amount of rainfall in three consecutive days reach at least 25mm and no dry spell of more than seven days occurs in the following thirty days. According to this study, the end of the rainy season is defined as that rainy day after which rain recorded during 20 days is less than 5mm. \cite{BenMohamed2002} has also found sea surface temperature anomalies at various locations and amount of rainfall in July, August and September to be significant for millet crops in Niger. The author has also considered the maximum air temperature in the hottest month (April) and the minimum air temperature in the coldest month (January) as possible predictors of crops in Niger, but he did not find them significant.

An important group of studies that have analysed relationship of yield and climate in Sub-Saharan Africa has utilised degree days (\citealt{SchulzeEtA1993,TingemEtAl2008,WalkerSchulze2008,TingemEtAl2009}) or number of days with temperature above certain level or within defined range (\citealt{Giannakopoulos2009,LauxEtAl}).

@@

### 2018-11-14 writing/draft1/manuscript18.tex
adding

\begin{abstract}
\noindent This paper contributes to better understanding of effects of drought on food security in Kenya which should lead to improving of early warning systems and food security. Our dataset consists of yearly panel of $47$ counties of Kenya describing the period of 1981-2017.
\\
...Applying the linear mixed effects models, we found that...
\end{abstract}

@@

Before beginning our research, we had to answer the following two questions: `How should we measure food security?' and `How should we measure drought?' Based on previous literature and data availability, there are three main approaches of how to measure food security. These include agricultural crop yields (see literature overview in Sections~\ref{Introduction} and further in this section), food prices or health or utility indicators. The relevant studies focused on food prices 
were used to estimate relationship between sea surface temperature and commodity prices in tropics

We focused on maize as this crop is the principal staple food in Kenya and it is grown at 90\% of all farms in Kenya \citep{FAO}.

\cite{brown2014book}: global, food prices and so \\
\cite{BROWN2015} Weather and international price shocks on food prices in the developing world.  Markets in Africa, South Asia, Latin America (regions that contain large segments of poor) \\
\cite{OxfamIDS}: GLOBE, CGE sub-Saharan Africa, central America north Africa, other low-income countries selected by OXFAM  \\
\cite{robinson2010}   Implications of food production and price shocks welfare in Ethiopia: a general equilibrium analysis CGE \\
\cite{Ubilava2018}: Estimate relationship between sea surface temperature and commodity prices using. tropics?? not sure about geographic focus entirely...

\textcolor{red}{We were also considering to use food prices or export and import prices/retail prices, however, there were some problems...maybe some references to those who also considered food prices??}
\vspace{2cm}

>>>>

\begin{abstract}
\noindent \textcolor{red}{..the full abstract to be written..} This paper contributes to better understanding of effects of drought on food security in Kenya which should lead to improving of early warning systems and food security. Our dataset consists of yearly panel of $47$ counties of Kenya describing the period of 1981-2017.
\\
...Applying the linear mixed effects models, we found that...
\end{abstract}

@@

Before beginning our research, we had to answer two essential % fundamental??
questions: `How should we measure food security?' and `How should we measure drought?' Taking into account previous literature and data availability, we considered three main approaches to measuring food security. These include agricultural crop yields (see literature overview in Sections~\ref{Introduction} and further in this section), food prices or health or utility indicators. Unfortunately we had to exclude the health and utility indicators because of data availability issues. \textcolor{red}{rozvest..} The relevant studies focused on food prices include \cite{Ubilava2018}, who has estimated relationship between sea surface temperature and commodity prices in tropics, \cite{robinson2010} who analysed implications of food production and price shocks welfare in Ethiopia or \cite{OxfamIDS} who have focused on sub-Saharan Africa, central America north Africa and other low-income countries selected by OXFAM. \cite{brown2014book} and \cite{BROWN2015} have developed a quantitative models to investigate relationship of  weather and international price shocks on food prices in the developing world. \textcolor{red}{or the disadvantage of using pricess is..}We, however, decided not to use food prices as a measure of food security because prices are usually strongly spatially autocorrelated across counties and this could cause the regression estimates to be inconsistent and inefficient unless taking into account the spatial autocorrelation structure. Modelling the structure of spatial autocorrelation would require developing and estimation of complicated structural equations and additional data which may not be available.  Therefore we opted for a \textcolor{red}{single} reduced form equation with crop yields as the dependent variable.
We focused on maize as this crop is the principal staple food in Kenya and it is grown at 90\% of all farms in Kenya \citep{FAO}.

### 2018-11-14 writing/draft1/manuscript19.tex
adding

\subsection{Measures of yield and climate}

Before beginning our research, we had to answer two essential % fundamental??
questions: `How should we measure food security?' and `How should we measure drought?' Taking into account previous literature and data availability, we considered three main approaches to measuring food security. These include agricultural crop yields (see literature overview in Sections~\ref{Introduction} and further in this section), food prices or health or utility indicators. Unfortunately we had to exclude the health and utility indicators because of data availability issues. \textcolor{red}{rozvest..} The relevant studies focused on food prices include \cite{Ubilava2018}, who has estimated relationship between sea surface temperature and commodity prices in tropics, \cite{robinson2010} who analysed implications of food production and price shocks welfare in Ethiopia or \cite{OxfamIDS} who have focused on sub-Saharan Africa, central America north Africa and other low-income countries selected by OXFAM. \cite{brown2014book} and \cite{BROWN2015} have developed a quantitative models to investigate relationship of  weather and international price shocks on food prices in the developing world. \textcolor{red}{or the disadvantage of using pricess is..}We, however, decided not to use food prices as a measure of food security because prices are usually strongly spatially autocorrelated across counties and this could cause the regression estimates to be inconsistent and inefficient unless taking into account the spatial autocorrelation structure. Modelling the structure of spatial autocorrelation would require developing and estimation of complicated structural equations and additional data which may not be available.  Therefore we opted for a \textcolor{red}{single} reduced form equation with crop yields as the dependent variable.
We focused on maize as this crop is the principal staple food in Kenya and it is grown at 90\% of all farms in Kenya \citep{FAO}.

>>>>

\subsection{Measures of food security and drought}

Before beginning our research, we had to answer two essential % fundamental??
questions: `How should we measure food security?' and `How should we measure drought?' Taking into account previous literature and data availability, we considered three main approaches to measuring food security. These include using food prices, health and utility indicators or agricultural crop yields as measures of food security. Unfortunately, we had to exclude the health and utility indicators because of data availability issues. Although some health and utility indicators, such as Mid Upper Arm Circumference (MUAC) of children under five, the Coping Strategies Index (\citealp{CSI}) or the early warning phases are available in the NDMA Early Warning Bulletins, they are only available since mid $2013$. Furthermore, the scales do not seem to be consistent over all counties and frequently the values are not present at all. The relevant studies focused on food prices include \cite{Ubilava2018}, who has estimated relationship between sea surface temperature and commodity prices in tropics, \cite{robinson2010} who analysed implications of food production and price shocks welfare in Ethiopia or \cite{OxfamIDS} who have focused on sub-Saharan Africa, central America north Africa and other low-income countries selected by OXFAM. \cite{brown2014book} and \cite{BROWN2015} have developed a quantitative models to investigate relationship of  weather and international price shocks on food prices in the developing world. \textcolor{red}{or the disadvantage of using pricess is..}We, however, decided not to use food prices as a measure of food security because prices are usually strongly spatially autocorrelated across counties and this could cause the regression estimates to be inconsistent and inefficient unless taking into account the spatial autocorrelation structure. Modelling the structure of spatial autocorrelation would require developing and estimation of complicated structural equations and additional data which may not be available.  Therefore we opted for a \textcolor{red}{single} reduced form equation with crop yields as the dependent variable.  We have provided a broad overview of literature concerned with agricultural yields in Sections~\ref{Introduction} and further in this section.
We focused on maize as this crop is the principal staple food in Kenya and it is grown at 90\% of all farms in Kenya \citep{FAO}.
